Drought frequency predicts life history strategies in \emph{Heliophila}

### 2019-03-08 manuscript/manuscript.tex
revisions

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Drought frequency predicts life history strategies in Heliophila},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn G. Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\shorttitle{Drought and life history}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{2} College of Agriculture, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{3} Institute for Environment and Society, Brown University, Providence, RI 02912, USA\\\textsuperscript{4} Biology Department, Pennsylvania State University, State College, PA 16802, USA}
\keywords{drought adaptation, life history evolution, remote sensing, phylogeography, herbaria records}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of stress events such as drought. Here we test this hypothesis
in \textit{Heliophila} (Brassicaceae), a diverse genus of flowering
plants native to Africa, by integrating 34 years of satellite-based
drought measurements with 2192 herbaria occurrence records. Consistent
with predictions from classic life history theory, we find that
perennial \textit{Heliophila} species occur in environments where
droughts are significantly less frequent compared to annuals. These
associations are predictive while controlling for phylogeny, lending
support to the hypothesis that drought related natural selection has
influenced the distributions of these strategies. Additionally, the
collection dates of annual and perennial species indicate that annuals
escape drought prone seasons during the seed phase of their life cycle.
Together, these findings provide empirical support for classic
hypotheses about the drivers of life history strategy in plants - that
perennials out compete annuals in environments with less frequent
drought and that annuals are adapted to environments with more frequent
drought by escaping drought prone seasons as seeds.

}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum in
angiosperms an important division exists, distinguishing annuals which
complete their seed to seed life cycle within a single calendar year
from perennials which can persist over multiple years. Annual plants
flower once, set seed, senesce, and then die, spending at least some
portion of the year as a seed, where they are relatively protected from
environmental stress. In contrast, perennial plants can continue
vegetative growth after reproduction and must survive conditions
experienced during all seasons. These represent fundamentally different
life history strategies, but the ecological factors that explain their
evolution and distributions remain empirically uresolved (Friedman \&
Rubin, 2015).\\
Classical theory predicts shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Stearns, 1992; Franco \&
Silvertown, 1996). In plants, this has been extended to the hypothesis
that annuality is adaptive when it allows plants to escape drought
(Schaffer \& Gadgil, 1975). Lack of water is perhaps the greatest threat
to survival during vegetative or reproductive growth and annuals can
remain dormant (and protected as a seed) during drought. Thus,
environments with greater seasonal drought frequency may select for
annual life histories that complete reproduction prior to drought prone
seasons. Conversely, environments with less frequent drought may select
for perennial species, which benefit from multiple bouts of reproduction
and competitive advantage by preventing recruitment of annual species
(Corbin \& D'Antonio, 2004). These predictions have been supported by
the observation of annuals in arid environments in \emph{Oryza perennis}
(Morishima \emph{et al.}, 1984) and \emph{Oenothera} (Evans \emph{et
al.}, 2005). Additionally, annual and perennial species of
\emph{Nemesia} were qualitatively associated with winter rather and
summer rainfall environments respectively (Datson \emph{et al.}, 2008)
and annual species of \emph{Scorzoneroides} were associated with
environments classified as unpredictable (Cruz-Mazo \emph{et al.},
2009). However, whether the history frequency of drought events indeed
predicts the distributions annual or perennial life history strategies
has yet to be tested.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Furthermore, if
annual species have adapted to escape drought prone seasons,
observations of growing annual species (i.e.~occurring in forms other
than seed) should be rare during drought prone seasons. Phylogenetic
relatedness can influence tests of associations between species' traits
and their environments (Felsenstein, 1985; Barrett \emph{et al.}, 1996),
and therefore we assessed the relationship between life history
distribution and drought frequency in a phylogenetic context.

@@

\hypertarget{availability}{%
\subsubsection{Availability}\label{availability}}

@@

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak \emph{et al.}, 2015). To quantify the frequency of drought
during different seasons across landscapes, we used the remotely sensed
Vegetative Health Index (VHI), which measures landscape scale reductions
in plant cover and temperature conditions characteristic of drought
(Kogan, 2001). Generated from data collected by NOAA AVHRR satellites
since 1981, the VHI combines Normalized Difference Vegetation Index
(NDVI) derived measures of vegetative stress (Vegetative Condition Index
- VCI) with temperature stress indicated by anomalies in thermal spectra
(Temperature Condition Index - TCI). The VHI of year \(y\) during week
\(w\) of \([1,52]\) at pixel \(i\) is derived from the following
equations, where \(n\) is the number of years observed.

\[VCI_{y,w,i} = 100\frac{NDVI_{y,w,i} - NDVI_{min}}{NDVI_{max} - NDVI_{min}}\]

\[TCI_{y,w,i} = 100\frac{T_{y,w,i} - T_{min}}{T_{max} - T_{min}}\]

\[VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})\] where
\(NDVI_{min} = min(NDVI_{1981,w,i}...NDVI_{1981+n,w,i})\) and
\(NDVI_{max} = max(NDVI_{1981,w,i}...NDVI_{1981+n,w,i})\) and
\(T_{min} = min(T_{1981,w,i}...T_{1981+n,w,i})\) and
\(T_{max} = max(T_{1981,w,i}...T_{1981+n,w,i})\)

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas \emph{et al.}, 2011; Kogan
\emph{et al.}, 2016). But they also present a new tool to study seasonal
patterns in the frequency of drought across environments and to test
hypotheses about the effect of drought on ecological and evolutionary
processes (Kerr \& Ostrovsky, 2003). As such, the VHI has been applied
recently to study drought related ecology of natural species and proven
useful for predicting intraspecific variation in drought tolerance
traits and genes (Mojica \emph{et al.}, 2016; Dittberner \emph{et al.},
2018; Monroe \emph{et al.}, 2018b). Here, we accessed VHI data at
\(16km^2\) resolution from 1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the \emph{Heliophila} species occurring there are considered
to make up the most diverse genus of the family Brassicaceae (Mummenhoff
\emph{et al.}, 2005; Mandáková \emph{et al.}, 2012). This genus includes
both perennial and annual species and this change in life history
strategy has likely arisen multiple independent times (Appel \&
Al-Shehbaz, 1997; Mummenhoff \emph{et al.}, 2005). Furthermore, the fine
scale climatic heterogeneity of Southern Africa is ideal for studying
the distribution of traits in relation to environmental parameters
(Sayre \emph{et al.}, 2013). We used life histories reported by
Mummenhoff \emph{et al.} (2005), grouping species with annual or
perennial life histories. Perenniality was defined based any form of
perennial life history (e.g., herbs, shrubs, mixed, etc). Because the
nature of species reported with mixed traits were unknown
(i.e.~plasticity vs.~genetic variation), we classified these species
here as perennial since they can maintain vegetative growth after
reproduction at least to some capacity.

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years (Thiers, 2016). Herbarium
specimens and their associated metadata have been used since the 1960s
to study species' geographical distributions (reviewed by Willis
\emph{et al.} (2017) and Lang \emph{et al.} (2018)). And as they become
digitized (Soltis, 2017), these collections have been used to study
relationships between trait distributions, geography, and climate (Davis
\emph{et al.}, 2015; Stropp \emph{et al.}, 2016; Wolf \emph{et al.},
2016; Václavı'k \emph{et al.}, 2017). To characterize the distributions
of annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (GBIF, 2018).

An alignment of ITS I and II sequences for \emph{Heliophila} species was
obtained from the authors of Mandáková \emph{et al.} (2012). Individual
ITS I and II sequences for \emph{Aethionema grandiflorum, Alliaria
petiolata, Cardamine matthioli, Chamira circaeoides}, and \emph{Rorippa
amphibia} were downloaded from Genbank.

@@

To characterize drought regimens across the distributions of annual and
perennial species of \emph{Heliophila}, we calculated drought during
different seasons at the location of observations for \emph{Heliophila}
records using the VHI. Specifically, we created global maps of the
frequencies of observing drought conditions (VHI\textless{}40, NOAA)
during the winter (quarter surrounding winter solstice), spring (quarter
surrounding spring equinox), summer (quarter surrounding summer
solstice) and fall (quarter surrounding fall equinox) from 1981 to 2015.
From these maps, the drought frequency during the winter, spring,
summer, and fall were extracted for the locations of all GBIF records.

@@

\hypertarget{phylogeny-construction}{%
\subsubsection{Phylogeny construction}\label{phylogeny-construction}}

Out group ( \emph{Aethionema grandiflorum, Alliaria petiolata, Cardamine
matthioli, Chamira circaeoides}, and \emph{Rorippa amphibia}) and
ingroup \emph{Heliophila} ITS I and II sequences were aaligned using
MAFFT (Katoh \emph{et al.}, 2002) with strategy G-INS-I, offset value
0.1, and all other options set as default. The \(GTR + \Gamma\) model of
nucleotide substitution was determined to best fit the data based on AIC
using jModelTest2 (Guindon \& Gascuel, 2003; Darriba \emph{et al.},
2012). A maximum clade credibility tree with branch lengths as relative
time was estimated by summarizing data from six runs of 100,000,000
generations of Bayesian Markov chain Monte Carlo conducted in BEAST 2
(Bouckaert \emph{et al.}, 2014). Model selection and phylogenetic
analyses were conducted through the CIPRES Science Gateway (Miller
\emph{et al.}, 2010).

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
total occurrence records of annual and perennial species by t-tests. To
account for variation in the number of occurrence records per species,
we next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. Because shared evolutionary history of
closely related species can lead to spurious associations between traits
and environments (Felsenstein, 1985), we tested for a relationship
between life history strategy and drought frequency while controlling
for phylogeny using phylogenetic logistic regression (Ives \& Garland,
2010).

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/phylogeny} \caption{Species and examples of herbaria specimens of
\emph{Heliophila} (a) Phylogeny and life history strategies of species
studied. Orange circles at branch tips mark annual species and blue
circles mark perennial species. Example herbaria specimens accessed via
GBIF of (a) \emph{H. minima}, (b) \emph{H. deserticola}, (c) \emph{H.
coronopifolia} and (d) \emph{H. ephemera}. Images (a,c,d) courtesy of
The Bavarian Natural History Collections (CC BY-SA 4.0) and (b) The
London Natural History Museum (CC BY 4.0). Links to images are found in
the supplement.}\label{fig:phylogeny}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of occurrence records of (a) annual and (b)
perennial \emph{Heliophila.} Drought frequency during the (c) winter,
(d) spring, (e) summer and (f) fall measured using the VHI. (g) Drought
frequencies during each season at the observation locations of annual
and perennial \emph{Heliophila} (t tests, ** = p \textless{} 0.01).}\label{fig:mapsboxplots}
\end{figure}

@@

There were clear visual differences between the distributions of the 960
annual and the 1232 perennial \emph{Heliophila} observation records (see
Figure \ref{fig:speciesmaps} for maps of individual species). While
annual species were generally found in the western regions of South
Africa and Namibia, primarily in the Cape Floristic Region and Succulent
Karoo (Figure \ref{fig:mapsboxplots}a), the occurrence of perennials
extended to the east coast of South Africa (Figure
\ref{fig:mapsboxplots}b).

@@

We thank members of the Sloan lab for generous feedback that improved
the quality of this work. This research was supported by NSF Award
1701918 and USDA-NIFA Award 2014-38420-21801 to JGM.

>>>>

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Drought frequency predicts life history strategies in Heliophila},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn G. Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records, Heliophia},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\shorttitle{Drought and life history}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{2} College of Agriculture, Colorado State University, Fort Collins, CO 80521, USA\\\textsuperscript{3} Institute for Environment and Society, Brown University, Providence, RI 02912, USA\\\textsuperscript{4} Biology Department, Pennsylvania State University, State College, PA 16802, USA}
\keywords{drought adaptation, life history evolution, remote sensing, phylogeography, herbaria records, Heliophia}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\abstract{
\begin{center}\textbf{Summary}\end{center} \par

Explaining variation in life history strategies is an enduring goal of
evolutionary biology and ecology. For plants, annual and perennial life
histories have long been thought to reflect adaptation to environments
that experience alternative drought regimes. Neverthless, empirical
support for this hypothesis from phylogeographic analyses remains
lacking.

\par

Here, we tested classic life history theory in \textit{Heliophila}
(Brassicaceae), a diverse genus of flowering plants native to Africa,
with a phylogentically constrained analysis integrating 34 years of
satellite-based drought detection with 2192 herbaria occurrence records.

\par

We found that annual and perennial life histories have evolved multiple
times in \textit{Heliophila} and that compared to annuals, perennial
species occur in environments where droughts are significantly less
frequent. We also report evidence that annual plants adapt to
predictable drought regimes by escaping drought prone seasons as seeds.

\par

These results provide compelling support for longstanding theoretical
predictions and a more nuanced understanding of life history evolution.
More broadly, this work outlines a scalable framework integrating
herbaria records and remote sensing to address outstanding questions in
evolutionary ecology. 
}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with some species
completing their life cycle in a number of weeks to others that live for
thousands of years (Brown, 1996). Along this continuum in angiosperms an
important division exists between annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons
but can also benefit from multiple bouts of reproduction and competative
advantages. These represent fundamentally different life history
strategies and predicting their occurance is important for community,
ecosystem, and agricultural ecology. However, the environmental factors
that explain their evolution and distributions remain empirically
uresolved (Friedman \& Rubin, 2015).

Classical theory predicts shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Stearns, 1992; Franco \&
Silvertown, 1996). Because lack of water is perhaps the greatest threat
to survival during vegetative or reproductive growth in plants, this
theory has been extended to the hypothesis that annuality is adaptive
when it allows plants to escape drought (Schaffer \& Gadgil, 1975).
Indeed, adaptation to drought, defined as episodes of increased aridity
causing plant sresss (Passioura, 1996), is often invoked as an
explanation for the success of annual species. And while a few cases are
cited where annuality appears to be more common in environments with
greater aridity (Stebbins Jr, 1952; Morishima \emph{et al.}, 1984), this
hypothesis has yet to be supported in a phylogenetic context. In one
previous study where this question was addressed phylogenetically,
(Evans \emph{et al.}, 2005) annuals were not found to be associated with
environments that experience more drought. This could be explained by
the relatively small number of species studied and the reliance on a
limited number of weather stations to characterize environments,
highlighting the need to develop more scalable methods to study trait
evolution phylogeographically.

It is also critical to consider another dimension of drought adaptation:
the expectation that annuality is most adaptive when droughts are not
only frequent but also predictable. That is, when the frequency of
drought is particularly high during certain seasons. Such predictability
is important for selection to lead to escape from those seasons which
are paricuarlary drought prone. While there has been at least one
example of annuality associated with environments qualitatively
classified as \enquote{predictable} in a general sense (Datson \emph{et
al.}, 2008), the seasonal predictability of drought experienced by
annuals has yet to be rigorously studied. As such, further empirical
work is needed to support the model of annuality as a mechanism of
drought adaptation via escape from drought prone seasons.

In addition to drought escape in annuals as a mechanism of adaptation to
frequent predictable droughts, droughts may be necessary for the success
of annuals more generally by acting as episodes of disturbance that
provide opportunites for annuals to establish and compete with sympatric
perennial species. Indeed, there is considerable evidence that
perennials dominate in envirnoments where distrurbance events are
infrequent (Rees \& Long, 1992; Corbin \& D'Antonio, 2004; Clary, 2012).
The resulting prediction from this hypothesis is that in the absence of
frequent drought, perenniality should evolve. However, little is known
about this component of life history evolution because previous work has
largely focused on the origins of annuality rather than perenniality
(Friedman \& Rubin, 2015).

Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Additionally, if
perenniality offers competative advantage in the absence of drought,
associations between life history and drought frequency should be
signficant in even if phylogenies are enriched for transitions from
annual to perenial life history strategy. Finally, if annual species
have adapted to escape predictably drought prone seasons, observations
of growing annual species (i.e.~occurring in forms other than seed)
should be rare during seasons when drought frequency is hightest.
Phylogenetic relatedness can influence tests of associations between
species' traits and their environments (Felsenstein, 1985; Barrett
\emph{et al.}, 1996). Therefore we assessed the relationship between
life history distribution and drought frequency in a phylogenetic
context.

@@

\hypertarget{data-availability}{%
\subsubsection{Data availability}\label{data-availability}}

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
Africa including the Cape Floristic and Succulent Karoo Regions. These
are among the most botanically diverse environments on Earth and the
\emph{Heliophila} species occurring there are considered to make up the
most diverse genus of the family Brassicaceae (Mummenhoff \emph{et al.},
2005; Mandáková \emph{et al.}, 2012). This genus includes both perennial
(n =21) and annual (n = 21) species and this change in life history
strategy has likely arisen multiple independent times (Appel \&
Al-Shehbaz, 1997; Mummenhoff \emph{et al.}, 2005). Furthermore, the fine
scale climatic heterogeneity of Southern Africa is ideal for studying
the distribution of traits in relation to environmental parameters
(Sayre \emph{et al.}, 2013). We used life histories reported by
Mummenhoff \emph{et al.} (2005), grouping species with annual or
perennial life histories. Perenniality was defined based any form of
perennial life history (e.g., herbs, shrubs, mixed, etc). Because the
nature of species reported with mixed traits were unknown
(i.e.~plasticity vs.~genetic variation), we classified these species
here as perennial since they can maintain vegetative growth after
reproduction at least to some capacity.

To characterize the distributions of annual and perennial
\emph{Heliophila} species, all (8670) records for the genus
\emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (GBIF, 2018). herbaria
records records such as these provide a rich data sources to
characterize the geographical distributions of species (Thiers, 2016;
Willis \emph{et al.}, 2017; Lang \emph{et al.}, 2018). And as they
become digitized (Soltis, 2017), these collections have been used to
study relationships between trait distributions, geography, and climate
(Davis \emph{et al.}, 2015; Stropp \emph{et al.}, 2016; Wolf \emph{et
al.}, 2016; Václavı'k \emph{et al.}, 2017).

An alignment of ITS I and II sequences for 21 annual and 21 perennial
\emph{Heliophila} species was obtained from the authors of Mandáková
\emph{et al.} (2012). Individual ITS I and II sequences for
\emph{Aethionema grandiflorum, Alliaria petiolata, Cardamine matthioli,
Chamira circaeoides}, and \emph{Rorippa amphibia} were downloaded from
Genbank.

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak \emph{et al.}, 2015). From an ecological perspective,
droughts are best defined as episodes of plant stress caused by elevated
aridity (Passioura, 1996). Thus remote sensing offers the additional
benefit for studying drought as an agent of natural selection because
plant stress caused by drought can be observed from space (Kogan,
1995a). The remotely sensed Vegetative Health Index (VHI) is one such
metric, which detects landscape scale reductions in plant cover and
temperature conditions characteristic of drought (Kogan, 2001).
Generated from data collected by NOAA AVHRR satellites since 1981, the
VHI is a composite index combining Normalized Difference Vegetation
Index (NDVI) derived quantification of vegetative stress (Vegetative
Condition Index - VCI) with temperature stress indicated by anomalies in
thermal spectra (Temperature Condition Index - TCI). These indicies were
developed to create an unbiased quantification of drought across
ecosystem types. The VHI of year \(y\) during week \(w\) of \([1,52]\)
at pixel \(i\) is derived from the following equations, where \(n\) is
the number of years observed.

\[VCI_{y,w,i} = 100\frac{NDVI_{y,w,i} - NDVI_{min,w,i}}{NDVI_{max,w,i} - NDVI_{min,w,i}}\]

Low values of VCI indicate episodes when plant cover is particularly low
for a given location during a given time of the year. Thus, it controls
for the location and season in quantifying plant stress.

\[TCI_{y,w,i} = 100\frac{T_{max,w,i}-T_{y,w,i}}{T_{max,w,i} - T_{min,w,i}}\]
Similarly, low TCI values indicate episdoes of high thermal stress shown
to be negatively correlated with precipition and soil moisture
(AghaKouchak \emph{et al.}, 2015).

\[VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})\] By combining VCI
and TCI, the VHI is distinguishes drought from other forms of vegetative
stress (Kogan, 1995b). The use of the VHI to detect drought has been
validated globally and across ecosystem types (AghaKouchak \emph{et
al.}, 2015), including in southern Africa, the focal region of this
study (Figure @ref:fig(mapsdroughtexamples)). To date, the VHI has most
often been applied for evaluating drought risk for agricultural research
(e.g., Rojas \emph{et al.}, 2011; Kogan \emph{et al.}, 2016). But it
also presents a tool to study seasonal patterns in the frequency of
drought across environments and to test hypotheses about the effect of
drought on ecological and evolutionary processes (Kerr \& Ostrovsky,
2003). As such, the VHI has been applied recently to study drought
related ecology of natural species and proven useful for predicting
intraspecific variation in drought tolerance traits and genes (Mojica
\emph{et al.}, 2016; Dittberner \emph{et al.}, 2018; Monroe \emph{et
al.}, 2018b). Here, we accessed VHI data at \(16km^2\) resolution from
1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species across their native range of
southern Africa.

@@

To characterize drought regimens across the distributions of annual and
perennial species of \emph{Heliophila}, we calculated drought during
different seasons at the location of observations for \emph{Heliophila}
records using the VHI. Specifically, we created global maps of the
frequencies of observing drought conditions between years
(VHI\textless{}40, NOAA) during the winter (quarter surrounding winter
solstice), spring (quarter surrounding spring equinox), summer (quarter
surrounding summer solstice) and fall (quarter surrounding fall equinox)
from 1981 to 2015. From these maps, the drought frequency during the
winter, spring, summer, and fall were extracted for the locations of all
GBIF records.

@@

\hypertarget{phylogeny-construction-and-ancestral-state-estimation}{%
\subsubsection{Phylogeny construction and ancestral state
estimation}\label{phylogeny-construction-and-ancestral-state-estimation}}

Outgroup ( \emph{Aethionema grandiflorum, Alliaria petiolata, Cardamine
matthioli, Chamira circaeoides}, and \emph{Rorippa amphibia}) and
ingroup \emph{Heliophila} ITS I and II sequences were aligned using
MAFFT (Katoh \emph{et al.}, 2002) with strategy G-INS-I, offset value
0.1, and all other options set as default. The \(GTR + \Gamma\) model of
nucleotide substitution was determined to best fit the data based on AIC
using jModelTest2 (Guindon \& Gascuel, 2003; Darriba \emph{et al.},
2012). A maximum clade credibility tree with branch lengths as relative
time was estimated by summarizing data from six runs of 100,000,000
generations of Bayesian Markov chain Monte Carlo conducted in BEAST 2
(Bouckaert \emph{et al.}, 2014). Model selection and phylogenetic
analyses were conducted through the CIPRES Science Gateway (Miller
\emph{et al.}, 2010). Ancestral state estimation was performed in R
using the package phytools (Revell, 2012) to genearte 10,000 stochastic
character maps simulated under an equal rates model of character
evolution for the trait life habit (annual or perennial).

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought across all occurrence records of between annual and
perennial species by t-tests. Independent tests were performed to
compare drought frequency between annuals and perennials during each
season. We also tested for a relationship between drought frequency and
life history, season, and their interaction by an analysis of variance
while including species as a random effect. To account for variation in
the number of occurrence records per species, we next calculated the
mean drought frequency during the winter, spring, summer and fall for
each species. Because shared evolutionary history of closely related
species can lead to spurious associations between traits and
environments (Felsenstein, 1985), we tested for a relationship between
life history strategy and drought frequency while controlling for
phylogeny using phylogenetic logistic regression (Ives \& Garland,
2010). This statistical approach is designed to control for the
confounding effects of geographic relationships explained by demographic
history rather than processes of natural selection.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/phylogeny} \caption{Species and examples of herbaria specimens of
\emph{Heliophila} (a) Phylogeny and life history strategies of species
studied. Orange circles at branch tips mark annual species and blue
circles mark perennial species. At internal nodes, pie charts indicate
the estimated posterior probability of being annual versus perennial.
Example herbaria specimens accessed via GBIF of (a) \emph{H. minima},
(b) \emph{H. deserticola}, (c) \emph{H. coronopifolia} and (d) \emph{H.
ephemera}. Images (a,c,d) courtesy of The Bavarian Natural History
Collections (CC BY-SA 4.0) and (b) The London Natural History Museum (CC
BY 4.0). Links to images are found in the supplement.}\label{fig:phylogeny}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of occurrence records of (a) annual and (b)
perennial \emph{Heliophila.} Drought frequency during the (c) winter,
(d) spring, (e) summer and (f) fall detected using the VHI. (g) Drought
frequencies during each season at the observation locations of annual
and perennial \emph{Heliophila} (t tests contrasting annuals and
perennials, ** = p \textless{} 0.01).}\label{fig:mapsboxplots}
\end{figure}

The \emph{Heliophila} phylogeny was consistent with previous studies
(Mummenhoff \emph{et al.}, 2005; Mandáková \emph{et al.}, 2012). Based
on 10,000 stochastic character maps simulated under an equal rates model
of character evolution in life history, an average of approximately
eight changes from annual to perennial and five changes from perennial
to annual are observed per stochastic character map (Figure
\ref{fig:phylogeny}a). These results suggest that the ancestral state of
\emph{Heliophila} was annual and that both charcter states have arisen
independently mutliple times.

@@

There were clear visual differences between the distributions of the 960
annual and the 1232 perennial \emph{Heliophila} observation records (see
Figure \ref{fig:speciesmaps} for maps of individual species). While
annual species were generally found in the western regions of South
Africa and Namibia, primarily in the Cape Floristic Region and Succulent
Karoo (Figure \ref{fig:mapsboxplots}a), the occurrence of perennials
extended to the southern and eastern coast of South Africa (Figure
\ref{fig:mapsboxplots}b).

@@

We thank Jesse Lasky, members of the Sloan lab at CSU and three
ananymous reviewers for generous feedback that improved the quality of
this work. This research was supported by NSF Award 1701918 and
USDA-NIFA Award 2014-38420-21801 to JGM.
